The Importance of Initial Conditions in Seasonal Predictions of Antarctic Sea Ice

### 2025-11-19 analysis/paper/access-ice.qmd
Adds Xiaobing comments

---
title: Evaluating the Importance of Initial Conditions for Antarctic Sea Ice Seasonal Predictability with a Fully Coupled Forecast Model
format:
  nature-pdf:
    # journal.cite-style is included in the tex file but ignored by pandoc if 
    # cite-method is not `natbib`.
    journal:
      cite-style: sn-mathphys-num
    # `citeproc` is the pandoc default. Set `cite-method: natbib` if required
    # to use the bst styles from the upstream template.
    cite-method: natbib
    keep-tex: true
    equal-margins: false
    classoption: referee,lineno
    fig-width: 5  
    fig-pos: "!htb"
header-includes: | 
   \usepackage[section]{placeins}
author:
  - name: Elio Campitelli
    affiliations:
      - name: Monash University, Kulin Nations, Clayton, Victoria, Australia.
        id: 1
        department: School of Earth, Atmosphere and Environment
        # address: 1737 Cambridge Street
        # city: Cambridge
        # state: MA
        # postal-code: 02138
      - name: Clayton, Kulin Nations, Victoria, Australia
        id: 2
        department: SARC Special Research Initiative for Securing Antarctica’s Environmental Future
        # address: 1 Oxford Street
        # city: Cambridge
        # state: MA
        # postal-code: 02138
    email: elio.campitelli@monash.edu
    attributes:
      corresponding: true
  - name: Ariaan Purich
    affiliations:
      - ref: 1
      - ref: 2
  - name: Julie Arblaster
    affiliations:
      - ref: 1
      - ref: 2
  - name: Eun-Pa Lim
    affiliations:
      - department: Research, Bureau of Meteorology, Australia
        id: 3
  - name: Matthew C. Wheeler
    affiliations:
      - ref: 3      
  - name: Phillip Reid
    affiliations:
      - ref: 3      
keywords: [sea ice, seasonal predictability, initial conditions, forecasting]
bibliography: references.bib

@@

The +S1 hindcast set is made up of nine members created by perturbing the atmospheric fields only with a random field perturbation [@hudson2017] and runs for 217 days for the period 1990–2012 initialised at the first of every month.
The +S2 hindcast set used in this study runs for the period 1981–2018. Each hindcast consists of nine ensemble members built from three consecutive three-member forecasts created in the same manner as +S1 members initialised at the first of every month and the two previous days and run for 279 days. 
We analyse the ensemble mean hindcasts unless otherwise specified.

@@

Forecasts errors are also compared with hypothetical forecasts based on the persistence of anomalies and on climatology. 
The persistence forecast is generated by extending the observed anomalies in time.

@@

```{r fig-hindcast-extent}
#| fig-cap: "Row a: Pan-Antarctic mean sea-ice extent for all hindcasts initialised on the first of each calendar month for +S1 (column 1; green) and +S2 (column 2; purple). Observed mean sea-ice extent in each corresponding hindcast period is shown in black. Row b: Mean differences between the forecast and the observed values. Circles represent the initial conditions at the start of forecasts (i.e., the first of every month), and triangles represent the mean values at the end of forecasts (i.e., the longest possible lead time)."
#| fig-height: 4

@@

  geom_hline(
    data = data.table(
      y = 0,
      measure = factor("Difference", levels = c("Forecast", "Difference"))
    ),
    aes(yintercept = 0)
  ) +
  geom_line(aes(
    group = interaction(forecast_time, time != time2),
    colour = model
  )) +
  geom_point(
    data = \(x) {
      x[
        mday(time) == 2,
        .SD[which.max(lag)],
        by = .(forecast_time, model, measure)
      ]
    },
    aes(fill = model),
    shape = 24,
    size = 2.2
  ) +
  geom_point(data = \(x) x[lag == 1], aes(fill = model), shape = 21, size = 2) +

@@

@fig-hindcast-extent shows mean sea-ice extent of the +S1 and +S2 hindcasts (row a) and their differences from mean sea-ice extent of +CDR (row b).
Mean extent at the first of every month is indicated with circles for the initial conditions and with triangles for the longest lead time possible for each model (between `r range_max_lag[model == "S2", min]` and `r range_max_lag[model == "S2", max]` days for +S2 and between `r range_max_lag[model == "S1", min]` and `r range_max_lag[model == "S1", max]` days for +S1).
At this long lead time, information of the initial conditions is essentially lost and the forecast reverts to each model's preferred equilibrium state.

@@

```{r forecast_example}
mlag <- 0
sample_date <- "2008-05-02"
forecast_date <- as.Date(sample_date)
month(forecast_date) <- month(sample_date) - mlag
mday(forecast_date) <- 1

@@

@@

As an example, @fig-forecast-example shows sea-ice concentration anomalies (top row) and sea-ice thickness and the difference between the two models (bottom row) for `r format(as.Date(sample_date), "%e %B %Y")` initialised one day prior; being that close to initialisation date, these are very approximately the initial conditions. 
+S1 sea-ice concentrations anomalies are very close to observations as expected from the system assimilating these data. 
+S2 sea-ice concentration anomalies, which are not assimilated, are not as close, but the large-scale pattern is aligned with observations.
The system simulates large positive anomalies in the Weddell and Ross Seas and slight negative anomalies in the Amundsen and Bellingshausen Seas.
The fact that +S2 can simulate this pattern without assimilating sea-ice data suggests that atmospheric and oceanic forcing were the dominant drivers.
However, the magnitude of the sea-ice anomalies is too big. It is plausible that this is due to the thinner ice simulated by +S2 (bottom row).

```{r fig-forecast-example}
#| fig.cap: +S1 and +S2 hindcasts for 2 May 2008 at one day lead time. Top row shows sea-ice concentration anomalies forecasted by each system and the observations. Bottom row shows forecasted sea-ice thickness and the difference between +S1 and +S2.
#| fig-height: 6

aice_forecast_example |>
  ggplot(aes(xgrid, ygrid)) +
  geom_contour_fill(
    aes(z = aice, fill = ..level..),
    breaks = AnchorBreaks(binwidth = 0.1, exclude = 0)
  ) +
  scale_fill_divergent_discretised(
    "Sea-ice concentration anomaly",
    low = trans_pink,
    high = trans_blue
  ) +
  geom_antarctica_fill +
  geomcoord_antarctica +
  facet_wrap(~model, labeller = labeller(model = labels_models)) +
  wide_legend +

  geom_contour_fill(
    data = \(x) {
      x |>
        dcast(time + xgrid + ygrid ~ model, value.var = "hi") |>
        _[, model := factor("Difference", levels = c("S1", "S2", "Difference"))]
    },
    aes(z = S1 - S2, fill = ..level..),
    breaks = c(-Inf, b, Inf)
  ) +

  scale_fill_divergent_discretised(
    "Sea-ice thickness difference",
    guide = guide_colorsteps(order = 0)
  ) +
  geom_antarctica_fill +
  geomcoord_antarctica +
  wide_legend +
  facet_wrap(~model) +
  plot_layout(ncol = 1)
```

@@

To assess +S2 forecasts in more detail, we compute error measures for all hindcasts started on the 1st of every month.

@@

@fig-rmse shows the mean RMSE of sea-ice concentration anomalies for +S1 and +S2 hindcasts compared against persistence and climatological forecasts used as a benchmark.
Due to errors in the initial conditions, it is expected that persistence forecasts would be better than the model forecasts at very short lead times, but that the persistence forecast errors would grow faster and may eventually surpass the model forecast errors.
The black line shows that the persistence forecast error indeed grows rapidly and reaches its maximum in about 30 days for most months except for February, when it grows much slower.
The +S1 forecast errors grow slower than persistence forecast errors and remain lower after less than 10 days on average.
The +S2 forecast error starts high in all months and is lower than the persistence forecast error after more than 15 days in most months except for forecast initialised in February, when it takes 80 days.

@@

range <- range(rmse_lon_mean_gdata2$skill)
binwidth <- 0.1
breaks <- AnchorBreaks(0, binwidth = binwidth)(range)
rmse_lon_plots <- lapply(c("S1", "S2"), \(x) {
  rmse_lon_mean_gdata2 |>
    _[version == x] |>
    # _[,
    #   skill := Smooth2D(lag, lon, skill, method = smooth_svd(0.05)),
    #   by = .(month, version)
    # ] |>
    ggplot(aes(lon, lag)) +
    geom_contour_fill(
      aes(z = skill, fill = after_stat(level)),
      breaks = breaks
    ) +
    geom_contour_tanaka(
      aes(z = skill),
      range = c(0.005, 0.05),
      breaks = breaks
    ) +

@@

```{r}
#| label: fig-rmse_lon-1
#| fig-cap: !expr glue::glue("RMSE skill score of +S1 forecasts with climatological forecast as reference computed on {uniqueN(rmse_lon_mean$lon)} meridional slices {diff(unique(rmse_lon_mean$lon))[1]}º wide as a function of lead time and longitude. Antarctica’s coastline is shown at the bottom of each panel for reference.")
#| fig-height: 7
rmse_lon_plots[[1]]
```

```{r}
#| label: fig-rmse_lon-2
#| fig-cap: Same as @fig-rmse_lon-1 but for +S2.
#| fig-height: 7
rmse_lon_plots[[2]]
```

@@

@@

Ensemble spread grows quickly even when perturbations are only implemented in the atmosphere component (in +S1), indicating that sea ice is indeed responding quickly to atmospheric perturbations. 
However, our analysis suggests that the atmosphere and ocean data assimilation implemented in +S2 is only effectively influencing sea-ice initial conditions from June to October, while the rest of the year, the sea-ice component runs virtually free, reverting to its biased equilibrium state.

@@

@@

>>>>

---
title: The Importance of Initial Conditions in Seasonal Predictions of Antarctic Sea Ice
format:
  nature-pdf:
    # journal.cite-style is included in the tex file but ignored by pandoc if 
    # cite-method is not `natbib`.
    journal:
      cite-style: sn-mathphys-num
    # `citeproc` is the pandoc default. Set `cite-method: natbib` if required
    # to use the bst styles from the upstream template.
    cite-method: natbib
    keep-tex: true
    equal-margins: false
    classoption: referee,lineno
    fig-width: 5  
    fig-pos: "!htb"
    fig-dpi: 300
header-includes: | 
   \usepackage[section]{placeins}
author:
  - name: Elio Campitelli
    affiliations:
      - name: Monash University, Kulin Nations, Clayton, Victoria, Australia.
        id: 1
        department: School of Earth, Atmosphere and Environment
        # address: 1737 Cambridge Street
        # city: Cambridge
        # state: MA
        # postal-code: 02138
      - name: Clayton, Kulin Nations, Victoria, Australia
        id: 2
        department: SARC Special Research Initiative for Securing Antarctica’s Environmental Future
        # address: 1 Oxford Street
        # city: Cambridge
        # state: MA
        # postal-code: 02138
    email: elio.campitelli@monash.edu
    attributes:
      corresponding: true
  - name: Ariaan Purich
    affiliations:
      - ref: 1
      - ref: 2
  - name: Julie Arblaster
    affiliations:
      - ref: 1
      - ref: 2
  - name: Eun-Pa Lim
    affiliations:
      - department: Research, Bureau of Meteorology, Australia
        id: 3
  - name: Matthew C. Wheeler
    affiliations:
      - ref: 3      
  - name: Phillip Reid
    affiliations:
      - ref: 3      
keywords: [sea ice, seasonal predictability, initial conditions, forecasting]
bibliography: references.bib

@@

The +S1 hindcast set is made up of nine members created by perturbing the atmospheric fields only with a random field perturbation [@hudson2017] and runs for 217 days for the period 1990–2012 initialised at the first of every month.
The +S2 hindcast set used in this study runs for the period 1981–2018.
Ensemble members are created in the same manner as +S1 members, however, due to computing cost limitations, only three members per forecast initialisation date were run for 279 days. 
Bigger ensambles were generated aggregating several three-member ensembles initialised on successive days [@wedd2022]. 
Here, we build a nine-member time-lagged ensemble from three consecutive three-member forecasts initialised at the first of every month and the two previous days and run for 279 days. 
We analyse the ensemble mean hindcasts unless otherwise specified.

@@

Forecasts errors are also compared with hypothetical forecasts based on the persistence of anomalies and on climatology. 
The persistence forecast is generated by extending the observed sea-ice concentration anomalies the day of the forecast initialisation and comparing it with the actual anomalies observed. 
The climatological forecast error is computed as the standard deviation of daily anomalies.

@@

```{r fig-hindcast-extent}
#| fig-cap: "Row a: Pan-Antarctic daily mean sea-ice extent for all hindcasts initialised on the first of each calendar month for +S1 (column 1; green) and +S2 (column 2; purple). Observed mean sea-ice extent in each corresponding hindcast period is shown in black. Row b: Mean differences between the forecast and the observed values. Circles represent the initial conditions at the start of forecasts (i.e., the first of every month), and triangles represent the mean values at the lead time corresponding to the maximum lead time in S1 (between 213 and 216 days, depending on the month)"
#| fig-height: 4

@@

  geom_hline(
    data = data.table(
      y = 0,
      measure = factor("Difference", levels = c("Forecast", "Difference"))
    ),
    aes(yintercept = 0)
  ) +
  geom_line(aes(
    group = interaction(forecast_time, time != time2),
    colour = model
  )) +
  # geom_point(
  #   data = \(x) {
  #     x[
  #       mday(time) == 2,
  #       .SD[which.max(lag)],
  #       by = .(forecast_time, model, measure)
  #     ]
  #   },
  #   aes(fill = model),
  #   shape = 24,
  #   size = 2.2
  # ) +

  geom_point(
    data = \(x) {
      x[
        mday(time) == 2,
        .SD[lag == max(lag[model == "S1"])],
        by = .(forecast_time, measure)
      ]
    },
    aes(fill = model),
    shape = 24,
    size = 2.2
  ) +

  geom_point(data = \(x) x[lag == 1], aes(fill = model), shape = 21, size = 2) +

@@

@fig-hindcast-extent shows mean sea-ice extent of the +S1 and +S2 hindcasts (row a) and their differences from mean sea-ice extent of +CDR (row b).
Mean extent at the first of every month is indicated with circles for the initial conditions and with triangles for the longest lead time possible for each model (between `r range_max_lag[model == "S2", min]` and `r range_max_lag[model == "S2", max]` days for +S2 and between `r range_max_lag[model == "S1", min]` and `r range_max_lag[model == "S1", max]` days for +S1).
At this long lead time, information of the initial conditions is essentially lost and the forecast reverts close to each model's preferred equilibrium state.

@@

```{r plot_forecast_example}

plot_forecast_example <- function(sample_date) {
  mlag <- 0
  forecast_date <- as.Date(sample_date)
  month(forecast_date) <- month(sample_date) - mlag
  mday(forecast_date) <- 1

@@

  aice_forecast_example |>
    _[, model := factor(model, labels = c("S1", "S2", which_dataset))] |>
    ggplot(aes(xgrid, ygrid)) +
    geom_contour_fill(
      aes(z = aice, fill = ..level..),
      breaks = AnchorBreaks(binwidth = 0.1, exclude = 0)
    ) +
    scale_fill_divergent_discretised(
      "Sea-ice concentration anomaly",
      low = trans_pink,
      high = trans_blue
    ) +
    geom_antarctica_fill +
    geomcoord_antarctica +
    facet_wrap(~model, labeller = labeller(model = labels_models)) +
    wide_legend +

    geom_contour_fill(
      data = \(x) {
        x |>
          dcast(time + xgrid + ygrid ~ model, value.var = "hi") |>
          _[,
            model := factor("Difference", levels = c("S1", "S2", "Difference"))
          ]
      },
      aes(z = S1 - S2, fill = ..level..),
      breaks = c(-Inf, b, Inf)
    ) +

    scale_fill_divergent_discretised(
      "Sea-ice thickness difference",
      guide = guide_colorsteps(order = 0)
    ) +
    geom_antarctica_fill +
    geomcoord_antarctica +
    wide_legend +
    facet_wrap(~model, labeller = labeller(model = labels_models)) +
    plot_layout(ncol = 1)
}

@@

As an example, @fig-forecast-example shows sea-ice concentration anomalies (top row) and sea-ice thickness and the difference between the two models (bottom row) for `r format(as.Date("2008-05-02"), "%e %B %Y")` initialised one day prior; being that close to initialisation date, these are very approximately the initial conditions. 
+S1 sea-ice concentrations anomalies are very close to observations as expected from the system assimilating these data. 
+S2 sea-ice concentration anomalies, which are not assimilated, are not as close, but the large-scale pattern is aligned with observations.
The system simulates large positive anomalies in the Weddell and Ross Seas and slight negative anomalies in the Amundsen and Bellingshausen Seas.
The fact that +S2 can simulate this pattern without assimilating sea-ice data suggests that atmospheric and oceanic forcing were the dominant drivers.
However, the magnitude of the sea-ice anomalies is too big. It is plausible that this is due to the thinner ice simulated by +S2 (bottom row).

```{r fig-forecast-example}
#| fig.cap: +S1 and +S2 hindcasts for 2 May 2008 at one day lead time. Top row shows sea-ice concentration anomalies forecasted by each system and the observations. Bottom row shows forecasted sea-ice thickness and the difference between +S1 and +S2.
#| fig-height: 6
plot_forecast_example(as.Date("2008-05-02"))
```

@@

@@

To assess +S2 forecasts in more detail, we compute error measures for all hindcasts started on the 1st of every month.
@fig-rmse shows the mean RMSE of sea-ice concentration anomalies for +S1 and +S2 hindcasts compared against persistence and climatological forecasts used as a benchmark.
Due to errors in the initial conditions, it is expected that persistence forecasts would be better than the model forecasts at very short lead times, but that the persistence forecast errors would grow faster and may eventually surpass the model forecast errors.
The black line shows that the persistence forecast error indeed grows rapidly and reaches its maximum in about 30 days for most months except for February, when it grows much slower.
The +S1 forecast errors grow slower than persistence forecast errors and remain lower after less than 10 days on average.
The +S2 forecast error starts high in all months and is lower than the persistence forecast error after more than 15 days in most months except for forecast initialised in February, when it takes 80 days.

@@

range <- range(rmse_lon_mean_gdata2$skill)
binwidth <- 0.1
breaks <- AnchorBreaks(0, binwidth = binwidth)(range)
rmse_lon_plots <- lapply(c("S1", "S2", "diff"), \(x) {
  if (x == "diff") {
    rmse_lon_mean_gdata2 <- rmse_lon_mean_gdata2 |>
      dcast(lag + lon + obs_dataset + month ~ version, value.var = "skill") |>
      _[, skill := S1 - S2]
    breaks <- NULL
  } else {
    rmse_lon_mean_gdata2 <- rmse_lon_mean_gdata2 |>
      _[version == x]
  }

  rmse_lon_mean_gdata2 |>
    ggplot(aes(lon, lag)) +
    geom_contour_fill(
      aes(z = skill, fill = after_stat(level)),
      breaks = breaks
    ) +
    geom_contour_tanaka(
      aes(z = skill),
      range = c(0.005, 0.05),
      breaks = breaks
    ) +

@@

```{r}
#| label: fig-rmse_lon-1
#| fig-cap: !expr glue::glue("RMSE skill score of +S1 forecasts with climatological forecast as reference computed on {uniqueN(rmse_lon_mean$lon)} meridional slices {diff(unique(rmse_lon_mean$lon))[1]}º wide as a function of lead time and longitude. Antarctica’s coastline is shown at the bottom of each panel for reference.")
#| fig-height: 7
#| dev: png
rmse_lon_plots[[1]]
```

```{r}
#| label: fig-rmse_lon-2
#| fig-cap: Same as @fig-rmse_lon-1 but for +S2.
#| fig-height: 7
#| dev: png
rmse_lon_plots[[2]]
```

@@

```{r}
#| label: fig-rmse_lon-3
#| fig-cap: Same as @fig-rmse_lon-1 but for the difference between +S1 and +S2.
#| fig-height: 7
#| dev: png
rmse_lon_plots[[3]]
```

Finally, @fig-rmse_lon-3 shows the difference in skill between +S1 and +S2. 
Large differences in skill indicate areas and months that are most affected by the data assimilation present in +S1.
Between January and March, which are the months in which +S1 is the most skillfuil (@fig-lead-time-window), most of the improvement compared with +S2 is present in the Ross and Weddell Sea. 
In April and May, the improvement seems more homogeneous.

@@

Ensemble spread grows quickly even when perturbations are only implemented in the atmosphere component (in +S1), indicating that sea ice is indeed responding quickly to atmospheric perturbations. 
However, our analysis suggests that the atmosphere and ocean data assimilation implemented in +S2 is only effectively influencing sea-ice initial conditions from June to October, while the rest of the year, the sea-ice component runs virtually free, reverting to its biased equilibrium state. 
@zhou2022 had previously evaluated sea-ice forecasts in +S2 and also highlighted the poor performance of this forecasting system attributed to the lack of good initial conditions.

@@

We thank the internal reviewers Bethan White and Xiaobing Zhou for their comments and feedback.

This work benefited by earlier unpublished work by Laura Davies, Phil Reid, Andrew G. Marshall.

This research was undertaken with the assistance of resources from the National Computational Infrastructure (NCI Australia), an NCRIS enabled capability supported by the Australian Government.

@@

```{r, fig-rmse_lon-3-osi, ref.label=I("fig-rmse_lon-3")}
